Locality and Effect Reflection

### 2025-07-08 paper.tex
The writing continues

\newcommand{\nat}{\mathbb{N}}
\newcommand{\glob}{\mathop{\Box}}
\newcommand{\cat}{\mathbb{C}}
\newcommand{\catobj}{C}
\newcommand{\op}{\mathtt{op}}
\newcommand{\catop}{\cat^{\op}}
\newcommand{\id}[1]{\mathrm{id}_{#1}}
\newcommand{\setc}{\mathrm{Set}}
\newcommand{\psh}{\widehat{\cat}}
\newcommand{\terminal}{\top}
\newcommand{\initial}{\bot}
\newcommand{\lawc}{\mathrm{Law}}
\newcommand{\lawcop}{\lawc^{\mathtt{op}}}
\newcommand{\pshlawcop}{\widehat{\lawcop}}
\newcommand{\freef}{Free}
\newcommand{\constfname}{\Omega}
\newcommand{\constf}[1]{\constfname #1}
\newcommand{\yonedaname}{y}
\newcommand{\yoneda}[1]{y(#1)}
\newcommand{\homset}[3]{\mathrm{Hom}_{#1}(#2, #3)}
\newcommand{\arrow}[2]{{#2}^{#1}}
\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
\newcommand{\reflectname}{\Uparrow}
\newcommand{\reflectraw}[1]{\mathop{\reflectname_{#1}}}
\newcommand{\reflect}[3]{\reflectraw{#1}(#2, #3)}
\newcommand{\reflectc}[2]{\mathop{\reflectname_{#1 #2}}}
\newcommand{\performraw}[2]{\mathop{\reflectname(#1(#2))}}
\newcommand{\perform}[5]{\performraw{#1}{#2}(#3, #4. #5)}
\newcommand{\reifyname}{\Downarrow}
\newcommand{\reifyraw}[1]{\mathop{\reifyname_{#1}}}
\newcommand{\reify}[3]{\reifyraw{#1}(#2.\,#3)}
\newcommand{\reifyc}[2]{\mathop{\reifyname_{#1 #2}}}
\newcommand{\outl}[1]{\mathop{\mathrm{out}_L} #1}
\newcommand{\outr}[1]{\mathop{\mathrm{out}_R} #1}
\newcommand{\mon}{\mathcal{T}}
\newcommand{\ind}[1]{W_{#1}}
\newcommand{\indintro}[4]{#2_{#1}(#3,\, #4)}
\newcommand{\retname}{\mathrm{ret}}
\newcommand{\ret}[2]{\mathop{\retname_{#1}}(#2)}
\newcommand{\indelim}[3]{\mathop{\mathrm{rec}_{#1}} #2 \,\{#3\}}
\newcommand{\indretcase}[2]{\retname(#1) \Rightarrow #2}
\newcommand{\indcase}[4]{#1(#2, #3) \Rightarrow #4}
\newcommand{\seq}{\,;\,}
\newcommand{\fst}{\mathop{\pi_1}}
\newcommand{\snd}{\mathop{\pi_2}}
\newcommand{\il}{\mathop{\iota_1}}
\newcommand{\ir}{\mathop{\iota_2}}
\newcommand{\inl}[1]{\mathop{\mathrm{in}_L} #1}
\newcommand{\inr}[1]{\mathop{\mathrm{in}_R} #1}
\newcommand{\case}[5]{\mathop{\mathrm{case}} #1 \,\{\,\inl{#2} \Rightarrow #3 \seq \inr{#4} \Rightarrow #5 \,\}}
\newcommand{\abs}[3]{\mathop{\lambda}(#1 \types #2).\,#3}
\newcommand{\app}[2]{#1\,#2}
\newcommand{\types}{\mathrel{:}}
\newcommand{\gtypes}{\mathrel{:_{\Box}}}
\newcommand{\cempty}{\varepsilon}
\newcommand{\ccons}[2]{#1;\,#2}
\newcommand{\lbind}[3]{\ccons{#1}{#2\types#3}}
\newcommand{\gbind}[3]{\ccons{#1}{#2\gtypes#3}}
\newcommand{\cdiv}[1]{#1/\Box}
\newcommand{\return}[1]{\mathop{\mathrm{return}} #1}
\newcommand{\letv}[3]{\mathop{\mathrm{let}} #1 = #2 \mathop{\mathrm{in}} #3}
\newcommand{\gbox}[1]{\mathop{\mathrm{box}} #1}
\newcommand{\gunbox}[3]{\mathop{\mathrm{let}} (\gbox #1) = #2 \mathop{\mathrm{in}} #3}
\newcommand{\grun}[1]{\mathop{\mathrm{run}}(#1)}
\newcommand{\sbst}[3]{#1[#2\backslash#3]}
\newcommand{\turnv}{\mathrel{\vdash_V}}
\newcommand{\turnc}{\mathrel{\vdash_C}}
\newcommand{\evmap}{\mathop{ev}}
\newcommand{\currymap}{\mathop{\lambda}}
\newcommand{\reducestov}{\rightsquigarrow_V}
\newcommand{\reducestoc}{\rightsquigarrow_C}
\newcommand{\mif}{\mathop{\mathrm{if}}}
\newcommand{\mthen}{\mathop{\mathrm{then}}}
\newcommand{\melse}{\mathop{\mathrm{else}}}
\newcommand{\embed}[1]{\mathopen{\lbag}#1\mathclose{\rbag}}

@@

\begin{itemize}
\item The ambient monad can be defined as a monad $\mon$ on such
  presheaves, which maps each algebraic theory to its free algebra. We use
  $\mon$ to define the semantics of computations.

@@

Terms in the free algebra of the sum of two effects
(i.e. \lstinline[style=oxcaml]{Term(Sum(L)(R)).t}) are isomorphic to a
recursive datatype that is built from the terms of one of the effects
and the operations of the other\cite{??}. We can definet this type as:
\begin{lstlisting}[style=oxcaml]
module Half_term (L : Op) (R : Op) : sig
  type 'a t =
    | Return : 'a @@ global -> 'a t
    | Op : 'r R.t @@ global * ('r -> Handler(L).t @ local -> 'a t) -> 'a t
end
\end{lstlisting}
and expose the isomorphism between
\lstinline[style=oxcaml]{Handler(L).t @ local -> Handler(R).t @ local -> 'a}
and \lstinline[style=oxcaml]{Handler(L).t @ local -> 'a Half_term(L)(R).t}:
\begin{lstlisting}[style=oxcaml]
module Half_reflection (L : Op) (R : Op) : sig
  module L_eff = Reflection(L)
  module R_eff = Reflection(R)
  module Sum_eff = Reflection(Sum(L)(R))
  module Proj = Project(L)(R)

@@

To this end, we will restrict the ambient monad to be an \emph{algebraic
  monad}: one that can be expressed as the free algebra of an algebraic
theory. Note that this defintion of algebraic monad is not precise:
different choices for the definition of ``algebraic theory'' give you
different meanings for ``algebraic monad''. We'll make our particular
choice over the course of this section.

@@

\begin{example}
  The free algebra of the theory of boolean state is equivalent
  to the type:
  \begin{lstlisting}[style=oxcaml]
  type 'a t =
    | Return of 'a
    | Get of unit * (bool -> 'a t)
    | Set of bool * (unit -> 'a t)
  \end{lstlisting}
  and it forms a monad:
  \begin{lstlisting}[style=oxcaml]
  let return x = Return x
  let rec bind x f =
    match x with
    | Return x -> f x
    | Get((), k) -> Get((), (fun s -> bind (k s) f))
    | Set(s, k) -> Set(s, (fun () -> bind (k ()) f))
  \end{lstlisting}  
\end{example}

@@

\subsection{Representing computations: The free algebra monad}

Our aim is to have a semantics where the meaning of types is indexed by
the ambient monad -- or more precisely by the algebraic theory whose
free algebra is the ambient monad. In particular, we would like the
interpretation of types of computations to be applications of the
ambient monad.

Given a Lawvere theory $L$, the underlying set of its free algebra forms
a monad $T_L$ on $\setc$. In addition to the unit of the monad, we can
construct elements of $T_L$ using the operations of $L$. For an
operation $op \types P \rightarrowtail A$ of a Lawvere theory $L$ and
some set $S$, there is a function
$op_{T_L} \types (P \times (A \rightarrow T_L(S))) \rightarrow T_L(S)$
that constructs terms in the free algebra using $op$.

$T_{\_}$ is functorial and we can use it to construct a monad $\mon$ on
$\pshlawcop$ such that:
\begin{equation*}
\mon(P)(L) = T_L(P(L))
\end{equation*}
We call $\mon$ the \emph{free algebra monad}, and we will use it as the
computation monad in our denotational semantics.
\begin{align*}
  &\sem{\Gamma \turnc v \types \tau} \mathrel{:}
    \homset{\pshlawcop}{\sem{\Gamma}}{\mon(\sem{\tau})} \\
  &\sem{\Gamma \turnc \return{v} \types \tau} =
    \eta_{\sem{\tau}} \circ \sem{\Gamma \turnv v \types \tau} \\
  &\sem{\Gamma \turnc \letv{x}{c_1}{c_2} \types \tau_2} =
    \mu_{\sem{\tau_2}}
    \circ \mon(\sem{\lbind{\Gamma}{x}{\tau_1} \turnc c_2 \types \tau_2})
    \circ t_{\sem{\Gamma}, \sem{\tau_1}}
    \circ \langle id_{\sem{\Gamma}}, \sem{\Gamma \turnc c_1 \types \tau_1} \rangle
\end{align*}
where $\eta$, $\mu$ and $t$ are the unit, multiplication and strength
respectively of $\mon$.

Categories of presheaves also have exponentials, which combine with the
free algebra monad to give us our semantics for function types.
\begin{align*}
  &\sem{\tau_1 \rightarrow \tau_2} = \arrow{\sem{\tau_1}}{\mon(\sem{\tau_2})} \\
  &\sem{\Gamma \turnv \abs{x}{\tau_1}{c} \types \tau_1 \rightarrow \tau_2} =
    \currymap(\sem{\lbind{\Gamma}{x}{\tau_1} \turnc c \types \tau_2})
     \\
  &\sem{\Gamma \turnc \app{v_1}{v_2} \types \tau_2} =
    \evmap
    \circ \langle \sem{\Gamma \turnv v_1 \types \tau_1 \rightarrow \tau_2},
                  \sem{\Gamma \turnv v_2 \types \tau_1} \rangle
\end{align*}
where $\evmap$ is the evaluation map and $\currymap$ the currying of
exponentials.

@@

For any category $\cat$, if $\cat$ has a terminal object $\terminal$, we
can define the global modality $\glob \types [\psh, \psh]$ such that:
$\glob(P)(L) = P(\terminal)$. In the case of $\pshlawcop$ this becomes:
\begin{equation*}
\glob(P)(L) = P(\initial)
\end{equation*}

Within our semantics, this amounts to restricting a type to only the
elements which do not depend on the ambient monad. This is very similar
to the purity comonad \todo{check name} in ?? \cite{??}. In particular,
\begin{equation*}
  \glob \mon(P) \quad = \quad T_{\initial} \circ \glob
  \cong \quad \glob P
\end{equation*}
as elements of the free algebra of the initial algebraic theory are just
elements of the carrier set.

For any non-empty Lawvere theory $L$, $\glob \yoneda{L}$ is the constant
presheave that always returns the empty set because there are no
morphisms from $L$ into $\initial$. In our denotational semantics this
amounts to all non-trivial global handler types being uninhabited.

\todo{fix this..}

Our denotational semantics in terms of $\pshlawcop$ is
straightforward. For types, the
translation $\sem{\tau} \mathrel{:} \pshlawcop$ is as follows:
\begin{mathpar}
\sem{\glob \tau} = \glob \sem{\tau} \and
\end{mathpar}
\begin{align*}
&\sem{\Gamma \turnv v \types \tau} \mathrel{:} \sem{\Gamma} \rightarrow \sem{\tau} \\
&\sem{\glob \Gamma \turnv \gbox{v} \types \glob \tau} =
  \glob \sem{\cdiv{\Gamma} \turnv v \types \tau} \circ \todo{weakening and globalizing}\\
&\sem{\Gamma \turnv \gunbox{x}{v_1}{v_2} \types \tau_2} =
  \sem{\gbind{\Gamma}{x}{\tau_1} \turnv v_2 \types \tau_2}
  \circ \langle id_{\sem{\Gamma}}, \sem{\Gamma \turnv v_1 \types \glob \tau_1} \rangle
&\sem{\glob \Gamma \turnv \grun{c} \types \glob \tau} =
  \epsilon \circ \todo{removing identity monad} \circ
  \glob \sem{\cdiv{\Gamma} \turnv v \types \tau} \circ \todo{weakening and globalizing}\\
\end{align*}
where $\epsilon$ is the counit of $\glob$.

@@

For an operation $op \types P \rightarrowtail A$ of a Lawvere theory $L$
and some set $S$, we can build a morphism:
\begin{equation*}
op_s \types \homset{\pshlawcop}{\yoneda{L}}{\mon{S}}
\end{equation*}

We use $\yoneda{L}$ as the semantics for our handler types.
\begin{align*}
&\sem{\yoneda{\Sigma}} = \yoneda{\sem{\Sigma}}\\
\end{align*}
where $\sem{\Sigma}$ is the Lawvere theory corresponding to the
signature $\Sigma$. If $\Sigma$ is:
\begin{equation*}
\{ \, op_1 \types \tau_{p_1} \rightarrowtail \tau_{a_1}; \ldots; op_n
\types \tau_{p_n} \rightarrowtail \tau_{a_n} \, \}
\end{equation*}
then $\sem{\Sigma}$ is the Lawvere theory for:
\begin{equation*}
\{ \, op_1 \types \sem{\tau_{p_1}}(\initial) \rightarrowtail \sem{\tau_{a_1}}(\initial); \ldots; op_n
\types \sem{\tau_{p_n}}(\initial) \rightarrowtail \sem{\tau_{a_n}}(\initial) \, \}
\end{equation*}
Note that by applying the types to the initial theory $\initial$ we
restrict them to their global elements: those that do not depend on the
current ambient monad.

For an operation $op \types P \rightarrowtail A$ of a Lawvere theory
$L$, we can define the following morphism of presheaves:

\todo{Write this}

that reflects the operation into the ambient monad. We use this as our
semantics of reflection.

\todo{fix this..}

\begin{align*}
&\sem{\Gamma \turnc \perform{v_1}{op}{v_2}{x}{c} \types \glob \tau} = \todo{fix}
  \evmap \langle \reflectc{\Sigma}{\tau} \circ \eta \circ
                   \sem{\Gamma \turnv v_1 \types \ind{\Sigma}(\glob \tau)},\,
                 \sem{\Gamma \turnv v_2 \types \yoneda{\Sigma}} \rangle \\
\end{align*}

Note that the Yoneda embedding preserves finite products, which for our
case means:
\begin{equation*}
\yoneda{L + L^{\prime}} \cong \yoneda{L} \times \yoneda{L^{\prime}}
\end{equation*}
which we use for the semantics of $\outl(v)$ and $\outr{v}$.

@@

Via the Yoneda lemma, we get the following:
\begin{equation*}
(\arrow{\yoneda{L}}{\mon(\glob A)})(L') \quad \cong \quad T_{(L' + L)}(A(\terminal))
\end{equation*}
which will form the basis of our language's isomorphisms between arrows
$Y \rightarrow \glob A$ and algebraic monads $M(\glob A)$.

Exponentials in categories of presheaves are defined such that:
\begin{equation*}
\arrow{P}{Q}(L) = \homset{\pshlawcop}{\yoneda{L} \times P}{Q}
\end{equation*}

\subsection{Inductive types and effect sums}

For any Lawvere theory $L$,
$\mu X.T_L(\sem{\tau} + F_\Sigma(X))$ is isomorphic to to
$T_{(L + \sem{\Sigma})}$ \cite{hyland2006combining}. That means that
$\mon(\sem{\ind{\Sigma}(\glob \tau)})$ is isomorphic to the presheaf
that maps $L$ to $T_{(L + \sem{\Sigma})}(\sem{\tau}(\terminal))$, which
in turn is isomorphic to $\sem{\yoneda{\Sigma} \rightarrow \glob \tau}$
by the result above. This isomorphism directly provides the semantics of
reification.

\todo{fix this..}

Our denotational semantics in terms of $\pshlawcop$ is
straightforward. We write $F_\Sigma$ for the endofunctor on $\setc$
for which $T_{\sem{\Sigma}}$ is the free monad. For types, the
translation $\sem{\tau} \mathrel{:} \pshlawcop$ is as follows:
\begin{mathpar}
\sem{\ind{\Sigma}(\tau)} = \mu X.(\sem{\tau} + F_\Sigma(\mon(X)))
\end{mathpar}
where $\mu X. F(X).$ represents the initial algebra of a suitable
endofunctor.

\begin{align*}
&\sem{\Gamma \turnv v \types \tau} \mathrel{:} \sem{\Gamma} \rightarrow \sem{\tau} \\
\end{align*}
\begin{align*}
&\sem{\Gamma \turnc c \types \tau} \mathrel{:} \sem{\Gamma} \rightarrow \mon(\sem{\tau}) \\
\end{align*}

\todo{fix this..}

Our denotational semantics in terms of $\pshlawcop$ is
straightforward.
\begin{align*}
&\sem{\Gamma \turnc \reify{\Sigma}{x}{c} \types \ind{\Sigma}(\glob \tau)} =
  \reifyc{\Sigma}{\tau} \circ \currymap(\sem{\lbind{\Gamma}{x}{\yoneda{\Sigma}} \turnc c \types \glob \tau}) \\
\end{align*}
$\reifyc{\Sigma}{}$ is the natural transformations in the isomorphism
from the previous section.

>>>>

\newcommand{\nat}{\mathbb{N}}
\newcommand{\glob}{\mathop{\Box}}
\newcommand{\cat}{\mathbb{C}}
\newcommand{\catobj}{C}
\newcommand{\op}{\mathtt{op}}
\newcommand{\catop}{\cat^{\op}}
\newcommand{\id}[1]{\mathrm{id}_{#1}}
\newcommand{\setc}{\mathrm{Set}}
\newcommand{\psh}{\widehat{\cat}}
\newcommand{\terminal}{\top}
\newcommand{\initial}{\bot}
\newcommand{\lawc}{\mathrm{Law}}
\newcommand{\lawcop}{\lawc^{\mathtt{op}}}
\newcommand{\pshlawcop}{\widehat{\lawcop}}
\newcommand{\freef}{Free}
\newcommand{\constfname}{\Omega}
\newcommand{\constf}[1]{\constfname #1}
\newcommand{\yonedaname}{y}
\newcommand{\yoneda}[1]{y(#1)}
\newcommand{\homset}[3]{\mathrm{Hom}_{#1}(#2, #3)}
\newcommand{\arrow}[2]{{#2}^{#1}}
\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
\newcommand{\reflectname}{\Uparrow}
\newcommand{\reflectraw}[1]{\mathop{\reflectname_{#1}}}
\newcommand{\reflect}[3]{\reflectraw{#1}(#2, #3)}
\newcommand{\reflectc}[2]{\mathop{\reflectname_{#1 #2}}}
\newcommand{\performraw}[2]{\mathop{\reflectname(#1(#2))}}
\newcommand{\perform}[5]{\performraw{#1}{#2}(#3, #4. #5)}
\newcommand{\reifyname}{\Downarrow}
\newcommand{\reifyraw}[1]{\mathop{\reifyname_{#1}}}
\newcommand{\reify}[3]{\reifyraw{#1}(#2.\,#3)}
\newcommand{\reifyc}[2]{\mathop{\reifyname_{#1 #2}}}
\newcommand{\outlname}{\mathrm{out}_L}
\newcommand{\outl}[1]{\mathop{\outlname} #1}
\newcommand{\outrname}{\mathrm{out}_R}
\newcommand{\outr}[1]{\mathop{\outrname} #1}
\newcommand{\mon}{\mathcal{T}}
\newcommand{\ind}[1]{W_{#1}}
\newcommand{\indintro}[4]{#2_{#1}(#3,\, #4)}
\newcommand{\retname}{\mathrm{ret}}
\newcommand{\ret}[2]{\mathop{\retname_{#1}}(#2)}
\newcommand{\indelim}[3]{\mathop{\mathrm{rec}_{#1}} #2 \,\{#3\}}
\newcommand{\indretcase}[2]{\retname(#1) \Rightarrow #2}
\newcommand{\indcase}[4]{#1(#2, #3) \Rightarrow #4}
\newcommand{\seq}{\,;\,}
\newcommand{\fst}{\mathop{\pi_1}}
\newcommand{\snd}{\mathop{\pi_2}}
\newcommand{\inl}[1]{\mathop{\mathrm{in}_L} #1}
\newcommand{\inr}[1]{\mathop{\mathrm{in}_R} #1}
\newcommand{\case}[5]{\mathop{\mathrm{case}} #1 \,\{\,\inl{#2} \Rightarrow #3 \seq \inr{#4} \Rightarrow #5 \,\}}
\newcommand{\abs}[3]{\mathop{\lambda}(#1 \types #2).\,#3}
\newcommand{\app}[2]{#1\,#2}
\newcommand{\types}{\mathrel{:}}
\newcommand{\gtypes}{\mathrel{:_{\Box}}}
\newcommand{\cempty}{\varepsilon}
\newcommand{\ccons}[2]{#1;\,#2}
\newcommand{\lbind}[3]{\ccons{#1}{#2\types#3}}
\newcommand{\gbind}[3]{\ccons{#1}{#2\gtypes#3}}
\newcommand{\cdiv}[1]{#1/\Box}
\newcommand{\return}[1]{\mathop{\mathrm{return}} #1}
\newcommand{\letv}[3]{\mathop{\mathrm{let}} #1 = #2 \mathop{\mathrm{in}} #3}
\newcommand{\gbox}[1]{\mathop{\mathrm{box}} #1}
\newcommand{\gunbox}[3]{\mathop{\mathrm{let}} (\gbox #1) = #2 \mathop{\mathrm{in}} #3}
\newcommand{\grun}[1]{\mathop{\mathrm{run}}(#1)}
\newcommand{\sbst}[3]{#1[#2\backslash#3]}
\newcommand{\turnv}{\mathrel{\vdash_V}}
\newcommand{\turnc}{\mathrel{\vdash_C}}
\newcommand{\evmap}{\mathop{ev}}
\newcommand{\currymap}{\mathop{\lambda}}
\newcommand{\reducestov}{\rightsquigarrow_V}
\newcommand{\reducestoc}{\rightsquigarrow_C}
\newcommand{\mif}{\mathop{\mathrm{if}}}
\newcommand{\mthen}{\mathop{\mathrm{then}}}
\newcommand{\melse}{\mathop{\mathrm{else}}}
\newcommand{\embed}[1]{\mathopen{\lbag}#1\mathclose{\rbag}}

@@

\begin{itemize}
\item The ambient monad can be defined as a monad $\mon$ on such
  presheaves, which maps each algebraic theory to its free algebras. We use
  $\mon$ to define the semantics of computations.

@@

Terms in the free algebras of the sum of two effects
(i.e. \lstinline[style=oxcaml]{Term(Sum(L)(R)).t}) are isomorphic to a
recursive datatype that is built from the terms of one of the effects
and the operations of the other\cite{??}. We can definet this type as:
\begin{lstlisting}[style=oxcaml]
module Half_term (L : Op) (R : Op) : sig
  type 'a t =
    | Return : 'a @@ global -> 'a t
    | Op : 'r R.t @@ global * ('r -> Handler(L).t @ local -> 'a t) -> 'a t
end
\end{lstlisting}
and expose the isomorphism between
\lstinline[style=oxcaml]{Handler(L).t @ local -> Handler(R).t @ local -> 'a}
and \lstinline[style=oxcaml]{Handler(L).t @ local -> 'a Half_term(L)(R).t}:
\begin{lstlisting}[style=oxcaml]
module Half_reflection (L : Op) (R : Op) : sig
  module L_eff = Reflection(L)
  module R_eff = Reflection(R)
  module Sum_eff = Reflection(Sum(L)(R))
  module Proj = Project(L)(R)

@@

To this end, we will restrict the ambient monad to be an \emph{algebraic
  monad}: one that can be expressed as the free algebras of an algebraic
theory. Note that this defintion of algebraic monad is not precise:
different choices for the definition of ``algebraic theory'' give you
different meanings for ``algebraic monad''. We'll make our particular
choice over the course of this section.

@@

\begin{example}
  The free algebras of the theory of boolean state are equivalent
  to the type:
  \begin{lstlisting}[style=oxcaml]
  type 'a t =
    | Return of 'a
    | Get of unit * (bool -> 'a t)
    | Set of bool * (unit -> 'a t)
  \end{lstlisting}
  and they form a monad:
  \begin{lstlisting}[style=oxcaml]
  let return x = Return x
  let rec bind x f =
    match x with
    | Return x -> f x
    | Get((), k) -> Get((), (fun s -> bind (k s) f))
    | Set(s, k) -> Set(s, (fun () -> bind (k ()) f))
  \end{lstlisting}  
\end{example}

@@

\subsection{Representing computations: The free algebras monad}

Our aim is to have a semantics where the meaning of types is indexed by
the ambient monad -- or more precisely by the algebraic theory whose
free algebras are the ambient monad. In particular, we would like the
interpretation of types of computations to be applications of the
ambient monad.

Given a Lawvere theory $L$, the underlying set of its free algebras form
a monad $T_L$ on $\setc$. In addition to the unit of the monad, we can
construct elements of $T_L$ using the operations of $L$. For an
operation $op \types P \rightarrowtail A$ of a Lawvere theory $L$ and
some set $S$, there is a function
$op_{T_L} \types (P \times (A \rightarrow T_L(S))) \rightarrow T_L(S)$
that constructs terms in the free algebra using $op$.

$T_{\_}$ is functorial and we can use it to construct a monad $\mon$ on
$\pshlawcop$ such that:
\begin{equation*}
\mon(P)(L) = T_L(P(L))
\end{equation*}
We call $\mon$ the \emph{free algebras monad}, and we will use it as the
computation monad in our denotational semantics.
\begin{align*}
  &\sem{\Gamma \turnc v \types \tau} \mathrel{:}
    \homset{\pshlawcop}{\sem{\Gamma}}{\mon(\sem{\tau})} \\
  &\sem{\Gamma \turnc \return{v} \types \tau} =
    \eta_{\sem{\tau}} \circ \sem{\Gamma \turnv v \types \tau} \\
  &\sem{\Gamma \turnc \letv{x}{c_1}{c_2} \types \tau_2} =
    \mu_{\sem{\tau_2}}
    \circ \mon(\sem{\lbind{\Gamma}{x}{\tau_1} \turnc c_2 \types \tau_2})
    \circ t_{\sem{\Gamma}, \sem{\tau_1}}
    \circ \langle id_{\sem{\Gamma}}, \sem{\Gamma \turnc c_1 \types \tau_1} \rangle
\end{align*}
where $\eta$, $\mu$ and $t$ are the unit, multiplication and strength
respectively of $\mon$.

Categories of presheaves also have exponentials, which combine with the
free algebras monad to give us our semantics for function types.
\begin{align*}
  &\sem{\tau_1 \rightarrow \tau_2} = \arrow{\sem{\tau_1}}{\mon(\sem{\tau_2})} \\
  &\sem{\Gamma \turnv \abs{x}{\tau_1}{c} \types \tau_1 \rightarrow \tau_2} =
    \currymap(\sem{\lbind{\Gamma}{x}{\tau_1} \turnc c \types \tau_2})
     \\
  &\sem{\Gamma \turnc \app{v_1}{v_2} \types \tau_2} =
    \evmap
    \circ \langle \sem{\Gamma \turnv v_1 \types \tau_1 \rightarrow \tau_2},
                  \sem{\Gamma \turnv v_2 \types \tau_1} \rangle
\end{align*}
where $\evmap$ is the evaluation map and $\currymap$ the currying of
exponentials.

@@

For any category $\cat$, if $\cat$ has a terminal object $\terminal$, we
can define the global modality $\glob \types [\psh, \psh]$ such that:
$\glob(P)(L) = P(\terminal)$. In the case of $\pshlawcop$ this becomes:
\begin{equation*}
\glob(P)(L) = P(\initial)
\end{equation*}
$\glob$ forms a comonad on $\pshlawcop$. $\glob$ also distributes over
products: $\glob (P \times Q) = (\glob P) \times (\glob Q)$

Within our semantics, this amounts to restricting a type to only the
elements which do not depend on the ambient monad. This is very similar
to the purity comonad in ??\cite{??}. In particular,
\begin{equation*}
  \glob \mon(P) \quad = \quad T_{\initial} \circ \glob P
  \quad \cong \quad \glob P
\end{equation*}
as elements of the free algebras of the initial algebraic theory are
just elements of their carrier sets.

We use the $\glob$ comonad as our semantics for $\glob$ types and
$\glob$ bindings.
\begin{align*}
&\sem{\glob \tau} = \glob \sem{\tau} \\
&\sem{\gbind{\Gamma}{x}{\tau}} = \sem{\Gamma} \times \glob \sem{\tau} \\
&\sem{\gbind{\Gamma}{x}{\tau} \turnv x \types \tau} = \epsilon \circ \snd \\
&\sem{\Gamma \turnv \gbox{v} \types \glob \tau} =
  \glob \sem{\cdiv{\Gamma} \turnv v \types \tau}
  \circ \eta \circ \cdiv{\Gamma} \\
&\sem{\Gamma \turnv \gunbox{x}{v_1}{v_2} \types \tau_2} =
  \sem{\gbind{\Gamma}{x}{\tau_1} \turnv v_2 \types \tau_2}
  \circ \langle id_{\sem{\Gamma}},
                \sem{\Gamma \turnv v_1 \types \glob \tau_1} \rangle \\
&\sem{\Gamma \turnv \grun{c} \types \glob \tau} =
  \rho_{\sem{\tau}}
  \circ \glob \sem{\cdiv{\Gamma} \turnc c \types \tau}
  \circ \eta \circ \cdiv{\Gamma}\\
\end{align*}
where $\epsilon$ and $\eta$ are the counit and comultiplication of
$\glob$, $\cdiv{\Gamma} \types \Gamma \rightarrow \cdiv{\Gamma}$ is the
morphism that projects the global bindings from $\Gamma$, and $\rho_P$
is the isomorphism from $\glob \mon(P)$ to $\glob P$.

@@

For an operation $op \types P \rightarrowtail A$ of a Lawvere theory $L$
and some set $S$, we can build a morphism:
\begin{align*}
&op_{\mon{S}} \types
  \homset{\pshlawcop}
         {\yoneda{L} \times \glob P \times \arrow{\glob A}{\mon{S}}}
         {\mon{S}} \\
&{(op_{\mon{S}})}_{L^\prime}
  = \lambda(y, p, k). (y(op))_{T_{L^\prime}}(p, \lambda a. k_{L^\prime}(id_{L^\prime}, a))
\end{align*}
which uses an element of $\yoneda{L}$ to map $op$ into the ambient monad
and then uses that to construct a term in the free algebras monad.

We use $\yoneda{L}$ as the semantics for our handler types, and
$op_{\mon{S}}$ as the semantics for reflection.
\begin{align*}
&\sem{\yoneda{\Sigma}} = \yoneda{\sem{\Sigma}} \\
&\sem{\Gamma \turnc \perform{v_1}{op}{v_2}{x}{c} \types \tau} = \\
&\quad op_{\mon{\sem{\tau}}}
       \circ \langle \sem{\Gamma \turnv v_1 \types \yoneda{\Sigma}},
                      \sem{\Gamma \turnv v_2 \types \tau_p},
                      \currymap(\sem{\lbind{\Gamma}{x}{\tau_a} \turnc c \types \tau})
             \rangle
\end{align*}
where $\sem{\Sigma}$ is the Lawvere theory corresponding to the
signature $\Sigma$. If $\Sigma$ is:
\begin{equation*}
  \{ \, op_1 \types \tau_{p_1} \rightarrowtail \tau_{a_1} \seq
  \ldots \seq
  op_n \types \tau_{p_n} \rightarrowtail \tau_{a_n} \, \}
\end{equation*}
then $\sem{\Sigma}$ is the Lawvere theory for:
\begin{equation*}
  \{ \, op_1 \types \sem{\tau_{p_1}}(\initial)
                    \rightarrowtail \sem{\tau_{a_1}}(\initial) \seq
  \ldots \seq
  op_n \types \sem{\tau_{p_n}}(\initial)
              \rightarrowtail \sem{\tau_{a_n}}(\initial) \, \}
\end{equation*}
Note that by applying the types to the initial theory $\initial$ we
restrict them to their global elements: those that do not depend on the
current ambient monad.

The Yoneda embedding preserves finite products, which for our case
means:
\begin{equation*}
\yoneda{L + L^{\prime}} \cong \yoneda{L} \times \yoneda{L^{\prime}}
\end{equation*}
which we use for the semantics of $\outlname$ and $\outrname$.

@@

We can represent computations in a specific algebraic monad using the
Yoneda lemma. The Yoneda lemma says that, for any $P \types \psh$ and
$\catobj \types \cat$, there is an isomorphism:
\begin{equation*}
  \homset{\psh}{\yoneda{\catobj}}{P} \cong P(\catobj)
\end{equation*}

Applying that to the composition of the free algebras monad and the
global modality we get:
\begin{equation*}
  \homset{\pshlawcop}{\yoneda{L}}{\mon{\glob \tau}} \cong \mon{\glob \tau}(L) = T_L(\tau(\initial))
\end{equation*}

In terms of our semantics, this means that computations of type
$\glob \tau$ with a single variable of type $\yoneda{\Sigma}$ are
isomorphic to values of the free algebra of $\sem{\Sigma}$ at
$\sem{\tau}(\initial)$.

\subsection{Effect sums and inductive types}

The essence of algebraic effect handlers is taking a computation in
$T_{L + \Sigma}(S)$ the free algebras of the sum of an algebraic theory
$L$ and an algebraic signature $\Sigma$, spliting out and handling
$\Sigma$, leaving a computation in only the algebraic theory
$T_{L}(S^\prime)$.

The key to this process is the following lemma
\cite{hyland2006combining}.
\begin{lemma}
\label{lemma:sums}
  For any Lawvere theory $L$ and algebraic signature $\Sigma$
  \begin{equation*}
  T_{L + \Sigma}(S) \cong \mu X.T_L(S + F_\Sigma(X))
  \end{equation*}
\end{lemma}
where $\mu X.F(X)$ is the intial algebra of a suitable endofunctor $F$,
and $F_\Sigma$ is the endofunctor on $\setc$ for which $T_{\Sigma}$ is
the free monad.

Effect reification works by exposing $\mu X.T_L(S + F_\Sigma(X))$ to the
user via an inductive type. In particular, given $\Sigma$ and a
presheave $P$, we can define the following initial algebra:
\begin{align*}
&\ind{\Sigma} \types [\pshlawcop, \pshlawcop] \\
&\ind{\Sigma}(P) = \mu X. (P + F_\Sigma \circ \mon{X})
\end{align*}
and observe the following variation on Lemma \ref{lemma:sums}.
\begin{equation*}
T_{L + \Sigma}(P(L)) \cong \mon{(\ind{\Sigma}(P))}(L)
\end{equation*}

We use these initial algebras as the denotational semantics of
$\ind{\Sigma}$:
\begin{mathpar}
\sem{\ind{\Sigma}(\tau)} = \ind{\sem{\Sigma}}(\sem{\tau})
\end{mathpar}

\subsection{Effect reification}

Exponentials in categories of presheaves are defined such that:
\begin{equation*}
\arrow{P}{Q}(\catobj) \cong \homset{\psh}{\yoneda{\catobj} \times P}{Q}
\end{equation*}
Applying this to $\arrow{\yoneda{L}}{\mon{(\glob P)}}$ and then applying
the Yoneda Lemma we get:
\begin{align*}
\arrow{\yoneda{L}}{\mon{(\glob P)}}(L')
  &\cong \homset{\pshlawcop}{\yoneda{L'} \times \yoneda{L}}{\mon{(\glob P)}} \\
  &\cong \homset{\pshlawcop}{\yoneda{L' + L}}{\mon{(\glob P)}} \\
  &\cong \mon{(\glob P)}(L' + L) \\
  &\cong T_{(L' + L)}(P(L'))
\end{align*}
